### 2017-03-19 paper/paper.tex
Wrote more

\hypersetup{
	colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=black
}

\usepackage[backend=biber,style=authoryear]{biblatex}
\usepackage[newfloat]{minted}
\usepackage{csquotes}

@@

\newcommand\tab[1][1cm]{\hspace*{#1}}
\SetupFloatingEnvironment{listing}{name=Code example}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\newcounter{codecounter}
\newenvironment{code}{\captionsetup{type=listing}}{}

@@

\addbibresource{sample.bib}
\usepackage{caption}

@@

The purpose of this essay is to evaluate differences and similarities in terms of election results in three different voting systems: Single Transferable Vote (STV), First-Past-The-Post (FPTP) and the Schulze Method. I will only be investigating and discussing differences and similarities in the results of the methods, not touching on how practically feasible the systems are in an actual election.

@@

Methodology here
All voting methods have been implemented in Typescript – mostly due to the author having previous experience with JavaScript. Typescript is a superset of JavaScript and compiles down to it. Typescript has some extra features compared to JavaScript but most importantly it has optional static type checking. Here is some example code in typescript:

@@

@@

The single transferable vote algorithm uses a tree structure to represent the state of the election. Every candidate is a top-level node with an attribute showing the number of current votes for the candidate. Each node has child nodes that show the next prefrences of the voters for that particular candidate, see figure \ref{Tree structure}. It is easy to manipulate branches via recursion. Merging and multiplying branches is used when transfering votes from one candidate node to another. A candidate node is a Typescript class. It can be found in the \texttt{stv/candidate.ts} file and has the following properties and functions:
%\lstinputlisting[language=JavaScript, label=some-code, caption=Some Code]{code/candidateNode.txt}

@@

With those functions, we can recursively manipulate branches and trees. In order to distribute the child nodes of an eliminated candidate, we just need to add each of those nodes to the corresponding top level nodes. In order to transfer surplus votes, we can multiply each child node with a factor and then distribute those nodes onto the top level nodes. The full source code for the tree manipuation functions can be found in \texttt{stv/tree.ts}.

The input into the algorithm is a set of ballots. A ballot consists of an ordered list of candidates according to prefrence. Every candidate is represented by an integer. The input data is converted into the tree structure via the \texttt{buildTree} function. It can also be found in the \texttt{stv/tree.ts} file.

The main procces of the program is found in the \texttt{stv/election.ts} file and runs the process defined in the psuedocode on page \pageref{Stv psuedocode}. The program takes the input defined above and a number representing the number of seats to elect as inputs. The process outputs an array of winning candidates as well as a log file.

@@

\begin{table}[H]
\centering
\begin{tabular}{l|c|c|c|c|c|}
\cline{2-6}
 & \multicolumn{1}{l|}{A} & \multicolumn{1}{l|}{B} & \multicolumn{1}{l|}{C} & \multicolumn{1}{l|}{D} & \multicolumn{1}{l|}{E} \\ \hline
\multicolumn{1}{|l|}{A} & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}17 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{DDFFDD}35 & \cellcolor[HTML]{DDFFDD}30 \\ \hline
\multicolumn{1}{|l|}{B} & \cellcolor[HTML]{DDFFDD}33 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}24 & \cellcolor[HTML]{DDFFDD}47 & \cellcolor[HTML]{DDFFDD}36 \\ \hline
\multicolumn{1}{|l|}{C} & \cellcolor[HTML]{DDFFDD}36 & \cellcolor[HTML]{DDFFDD}26 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{DDFFDD}40 & \cellcolor[HTML]{DDFFDD}42 \\ \hline
\multicolumn{1}{|l|}{D} & \cellcolor[HTML]{FFDDDD}15 & \cellcolor[HTML]{FFDDDD}3 & \cellcolor[HTML]{FFDDDD}10 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}18 \\ \hline
\multicolumn{1}{|l|}{E} & \cellcolor[HTML]{FFDDDD}20 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{FFDDDD}8 & \cellcolor[HTML]{DDFFDD}32 & \cellcolor[HTML]{9B9B9B} \\ \hline
\end{tabular}
\caption{Pairwise comparison matrix used in the Schulze Mehtod. For example, 17 ballots prefer $A>B$ and 33 ballots prefer $B>A$.}
\label{Pairwise comparison matrix}
\end{table}
\subsubsection{Justification}
The Shulze Method is used by multiple organisations around the world. It has been used by various software organisations such as The Wikimedia Foundation, The Debian Project and Ubuntu. It is also used by political parties such as the Pirate Party in Sweden and various other countries. The process also differs greatly in both method and implementation from the two other voting methods used in this paper.
\subsubsection{Pseudocode}
\label{Schulze psuedocode}
let $d[i,j]$ be the number of ballots that prefer candidate $i$ to candidate $j$\\
let $p[i,j]$ be the strength of the strongest path from candidate $i$ to candidate $j$\\
for $i$ from 1 to $C$\\
\tab for $j$ form 1 to $C$\\
\tab\tab if ($i \ne j$)\\
\tab\tab\tab if ($d[i,j] > d[j,i]$)\\
\tab\tab\tab\tab $p[i,j] \coloneqq d[i,j]$\\
\tab\tab\tab else\\
\tab\tab\tab\tab $p[i,j] \coloneqq 0$\\
\tab\tab\tab endif \\
\tab\tab endif \\
\tab endfor \\
endfor\\
for $i$ from 1 to $C$\\
\tab for $j$ form 1 to $C$\\
\tab\tab if ($i \ne j$)\\
\tab\tab\tab for $k$ from 1 to $C$\\
\tab\tab\tab\tab if ($i \ne k \text{ and } j \ne k$)\\
\tab\tab\tab\tab\tab $p[j,k] \coloneqq \text{max }(p[j,k], \text{ min }(p[j,k], p[i,k]))$\\
\tab\tab\tab\tab endif \\
\tab\tab\tab endfor \\
\tab\tab endif \\
\tab endfor \\
endfor\\
\subsubsection{Implementation}
The Shulze Method is implemented within the file \texttt{schulze/index.ts}. It takes two inputs: a list of ballots and the number of seats to be elected. A ballot is an ordered list of candidates. The output of the program is a list of winners.
\section{Evaluation of methods}
\subsection{Sample I}
\subsection{Sample II}
\subsection{Sample III}

@@

\begin{figure}
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar stacked,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			cycle list = {black!80,black!50,black!20},
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {Elections won},
			xlabel = {Party},
			ymajorgrids = true,
			every axis plot/.append style={fill,draw=none,no markers},
      ymin = 0
      ]
			\addplot coordinates
			{(0, 1)(1, 1)(2, 1)(3, 0)(4, 0)(5, 0)(6, 0)};
			\addplot coordinates
			{(0, 1)(1, 1)(2, 1)(3, 0)(4, 0)(5, 0)(6, 0)};
			\addplot coordinates
			{(0, 0)(1, 1)(2, 1)(3, 0)(4, 1)(5, 0)(6, 0)};
			\legend{FPTP, STV, Schulze}
		\end{axis}
	\end{tikzpicture}
\end{figure}

\begin{figure}
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {First prefrence votes},
			xlabel = {Party},
			ymajorgrids = true,
			every axis plot/.append style={fill,draw=none,no markers},
      ymin = 0
      ]
			\addplot coordinates
			{(0, 17)(1, 36)(2, 32)(4, 9)(5, 5)(6, 1)};
		\end{axis}
	\end{tikzpicture}
\end{figure}

>>>>

W
\hypersetup{
	colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=black
}

\usepackage[backend=biber,style=authoryear]{biblatex}
\usepackage[newfloat]{minted}
\usepackage{csquotes}
\usepackage{caption}

@@

\newcommand\tab[1][1cm]{\hspace*{#1}}
\SetupFloatingEnvironment{listing}{name=Code example}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\newenvironment{code}{\captionsetup{type=listing}}{}

@@

\newcommand\Code[1]{\texttt{#1}}

\addbibresource{sample.bib}

@@

The purpose of this essay is to evaluate differences and similarities in terms of election results in three different voting systems: Single Transferable Vote (STV), First-Past-The-Post (FPTP) and the Schulze Method. This paper will only be examining and discussing differences and similarities in the results of the methods, not touching on how practically feasible the methods might be in actual election.

@@

All voting methods have been implemented in Typescript – mostly due to the author having previous experience with JavaScript. Typescript is a superset of JavaScript with multiple extra feautures. Most importantly Typescript has optional static type checking. Here is some example code in typescript:

@@

In order to test the methods and their implementations, test data is needed. There are cases where extensive election data is published, such as in maltese elections, but full ballot data, which is needed for the implementations has not been found. Instead, the implementations will be tested with data generated from a simple computer algorithm. The algorithm used in this paper is highly primitive, as modeling prefrences within a population is far beyond this paper. The full ballot generation program can be found in the \Code{/generator/} folder. The goal of the program is to produce ballots that rank every candidate in order of prefrence (see figure \ref{STV ballot} on page \pageref{STV ballot}). The program represents each voter in a population as belonging to a certain ideology. Each ideology has has a list of weighted prefrences indicating how a voter for this ideology is likely to vote. The weighed prefrences are generated via an abstraction. A single config object is used to create all ideologies.
\codeblock{code/generator/ideologyConfig.ts}{code:ideology config}{Configuration object for creating ideologies}
The program then works via 4 major steps:
\begin{enumerate}
	\item It relates each candidate to an ideology based on the size of the ideology. A larger ideology has more candidates.
	\item It assigns a base size for each candidate. Each ideology has a different spread of votes generated by a negative exponential function with a different constant. The size of each candidate is then multiplied with the size of their ideology and normalized. This gives a base size of all candidates.
	\item Based on the base size of each candidate, a probability-list of voting for each candidate is created for every ideology. An ideology's probability-list represents how a voter aligned the ideology is expected to vote.
	\item It creates a certain number of individual ballots. Each ballot is given an ideology and creates its own list prefrences based via weighted random number generaton according to the probability-list of the assigned ideology.
\end{enumerate}
The output of the program is a set of ballots, where each ballot is an ordered set of prefrences. The implementation is highly arbitrary and all constants in the program even more so. However, expanding on the generator program would be beyond the scope of this project. The program works in that it is possible to control variables to test for different election scenarios.
\subsection{Constructing test scenarios}
Constructing test scenarios is difficult task. Due to there being so many variables both in real elections and the simulated elections in the generator program, it makes it practically inpossible to control all variables. Instead, 3 arbitrary test scenarios will be used. The election will be simulated 1000 times for each scenario and the election data will then run through the different voting methods and the results will be compared. The simulation scenarios are arbitrary but should allow for comparison of the different voting methods. All simulations will have 3 ideologies and 500 voters.
\subsubsection{Scenario 1}
Scenario 1 is this:
\codeblock{code/generator/scenario1.ts}{code:scenario 1}{Scenario 1}

@@

The single transferable vote algorithm uses a tree structure to represent the state of the election. Every candidate is a top-level node with an attribute showing the number of current votes for the candidate. Each node has child nodes representing the next prefrences of the voters for that particular candidate, see figure \ref{Tree structure}. It is easy to manipulate branches via recursion. Merging and multiplying branches is used when transfering votes from one candidate node to another. A candidate node is a Typescript class. It can be found in the \Code{stv/candidate.ts} file and has the following properties and functions:
\begin{figure}

@@

\codeblock{code/stv/candidateNode.ts}{code:stv candidate node}{The CandidateNode class}
With the \Code{add()} and \Code{multiply()} functions it is possible to recursively modify branches. There are two different manipulations that need to be preformed on the tree: tansfering surplus votes and eliminating candidates. Both functions rely on the \Code{distribute()} function wich is used to distribute a set of nodes onto the tree.
\codeblock{code/stv/distribute.ts}{code:stv distribute function}{The distribute function}
In order to eliminate and transfer surplus votes from any candidate node $A$, the children of $A$ are passed into the distribute function. In the case of transfering surplus votes, the votes for each child node is first multiplied by $\frac{\text{surplus votes}}{\text{total votes}}$ as outlined in the psuedocode in section \ref{Stv psuedocode}. The full source code for funcions that manipulate the tree structure can be found in \Code{/stv/tree.ts}.

The main procces of the program is found in the \Code{stv/election.ts} file and runs the process defined in the psuedocode in section \ref{Stv psuedocode} on page \pageref{Stv psuedocode}.

@@

\begin{table}[H]
\centering
\begin{tabular}{l|c|c|c|c|c|}
\cline{2-6}
 & \multicolumn{1}{l|}{A} & \multicolumn{1}{l|}{B} & \multicolumn{1}{l|}{C} & \multicolumn{1}{l|}{D} & \multicolumn{1}{l|}{E} \\ \hline
\multicolumn{1}{|l|}{A} & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}17 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{DDFFDD}35 & \cellcolor[HTML]{DDFFDD}30 \\ \hline
\multicolumn{1}{|l|}{B} & \cellcolor[HTML]{DDFFDD}33 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}24 & \cellcolor[HTML]{DDFFDD}47 & \cellcolor[HTML]{DDFFDD}36 \\ \hline
\multicolumn{1}{|l|}{C} & \cellcolor[HTML]{DDFFDD}36 & \cellcolor[HTML]{DDFFDD}26 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{DDFFDD}40 & \cellcolor[HTML]{DDFFDD}42 \\ \hline
\multicolumn{1}{|l|}{D} & \cellcolor[HTML]{FFDDDD}15 & \cellcolor[HTML]{FFDDDD}3 & \cellcolor[HTML]{FFDDDD}10 & \cellcolor[HTML]{9B9B9B} & \cellcolor[HTML]{FFDDDD}18 \\ \hline
\multicolumn{1}{|l|}{E} & \cellcolor[HTML]{FFDDDD}20 & \cellcolor[HTML]{FFDDDD}14 & \cellcolor[HTML]{FFDDDD}8 & \cellcolor[HTML]{DDFFDD}32 & \cellcolor[HTML]{9B9B9B} \\ \hline
\end{tabular}
\caption{Pairwise comparison matrix used in the Schulze Mehtod. For example, 17 ballots prefer $A>B$ and 33 ballots prefer $B>A$.}
\label{Pairwise comparison matrix}
\end{table}
\subsubsection{Justification}
The Shulze Method is used by multiple organisations around the world. It has been used by various software organisations such as The Wikimedia Foundation, The Debian Project and Ubuntu. It is also used by political parties such as the Pirate Party in Sweden and various other countries. The process also differs greatly in both method and implementation from the two other voting methods used in this paper.
\subsubsection{Pseudocode}
\label{Schulze psuedocode}
let $d[i,j]$ be the number of ballots that prefer candidate $i$ to candidate $j$\\
let $p[i,j]$ be the strength of the strongest path from candidate $i$ to candidate $j$\\
for $i$ from 1 to $C$\\
\tab for $j$ form 1 to $C$\\
\tab\tab if ($i \ne j$)\\
\tab\tab\tab if ($d[i,j] > d[j,i]$)\\
\tab\tab\tab\tab $p[i,j] \coloneqq d[i,j]$\\
\tab\tab\tab else\\
\tab\tab\tab\tab $p[i,j] \coloneqq 0$\\
\tab\tab\tab endif \\
\tab\tab endif \\
\tab endfor \\
endfor\\
for $i$ from 1 to $C$\\
\tab for $j$ form 1 to $C$\\
\tab\tab if ($i \ne j$)\\
\tab\tab\tab for $k$ from 1 to $C$\\
\tab\tab\tab\tab if ($i \ne k \text{ and } j \ne k$)\\
\tab\tab\tab\tab\tab $p[j,k] \coloneqq \text{max }(p[j,k], \text{ min }(p[j,k], p[i,k]))$\\
\tab\tab\tab\tab endif \\
\tab\tab\tab endfor \\
\tab\tab endif \\
\tab endfor \\
endfor\\
\subsubsection{Implementation}
The Shulze Method is implemented within the file \Code{schulze/index.ts}. It takes two inputs: a list of ballots and the number of seats to be elected. A ballot is an ordered list of candidates. The output of the program is a list of winners. The program first creates the matrices for both the prefrences and paths between candidates.
\codeblock{code/schulze/maps.ts}{code:schulze maps}{Data representation}
The program then runs the algorithm outlined in the psuedocode in section \ref{Schulze psuedocode}.
\codeblock{code/schulze/algorithm.ts}{code:schulze algorithm}{Main algorithm}
\section{Evaluation of methods}
\subsection{Sample I}
\subsection{Sample II}
\subsection{Sample III}

@@

\begin{figure}
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {Elections won},
			xlabel = {Candidate},
			ymajorgrids = true,
			every axis plot/.append style={draw=none,fill,no markers},
      ymin = 0,
			symbolic x coords={0,1,2,3,4,5,6,7,8},
			bar width=0.25cm
      ]
			\addplot coordinates
{(0, 563)(1, 337)(2, 100)(3, 0)(4, 0)(5, 0)(6, 0)(7, 0)};
\addplot coordinates
{(0, 323)(1, 203)(2, 474)(3, 0)(4, 0)(5, 0)(6, 0)(7, 0)};
\addplot coordinates
{(0, 333)(1, 255)(2, 412)(3, 0)(4, 0)(5, 0)(6, 0)(7, 0)};

			\legend{FPTP, STV, Schulze}
		\end{axis}
	\end{tikzpicture}
\end{figure}

\begin{figure}
	\centering
	\begin{tikzpicture}
		\begin{axis}[
			ybar,
			width = 0.9\textwidth,
			height = 0.5\textwidth,
			legend style={at={(0.5,-0.25)},
			anchor=north,legend columns=-1},
      x tick style = transparent,
			ylabel = {First prefrence votes},
			xlabel = {Candidate},
			ymajorgrids = true,
			every axis plot/.append style={fill,draw=none,no markers},
      ymin = 0
      ]
			\addplot coordinates
			{(0, 17)(1, 36)(2, 32)(4, 9)(5, 5)(6, 1)};
		\end{axis}
	\end{tikzpicture}
\end{figure}
